---
title: "Demo Code for estimation of BLUPs and Reliability using ASREML with PYT 2024"
author: "Vishnu Ramasubramanian"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document explains the R code for fitting **random-effects ASReml models** and estimating BLUPs and reliability (heritability) metrics across multiple traits and locations in plant breeding trials.

The analysis includes:

- random-effects models for genotypes (Strains)
- spatial fixed-effects models (Row and Range)
- calculation of BLUPs
- estimation of reliability (heritability) values
- comparison of models using AIC

---

# Part 1: Random-Effects Models per Location

## Model Setup

Lists are initialized to store models, predictions, and heritability:

```{r, eval=FALSE}

 modelRnd_Loc_List <- list()
 BLUPs_Loc_List <- list()
 BLUPs_LocTst_List <- list()
 wald.modelRnd_Loc_List <- list()
 i2_Loc_List <- list()


i2_Loc_Trts_List <- list()
i2_Loc_RR_Trts_List <- list()
```

---

## Loop Over Traits

For each trait: All the steps are implemented for each of the traits. However this markdown describes the analysis for one trait and one location as for loop is compatible with markdown description

```{r, eval=FALSE}
for(trait in trait.list){
  # model fitting steps
  
}

  nTrt <- 1

  trait <- trait.list[nTrt]
  trtInd <- which(!is.na(data.regular[,trait]))
  data.regular.trait <- data.regular[trtInd,]
  Locations <- levels(factor((data.regular.trait$Location)))
  
 # for(nLoc in 1:length(Locations)){ }  
  
    loc <- Locations[nLoc]
    
    data.loc <- data.regular.trait[data.regular.trait$Location ==loc,]
      
    data.loc$Strain <- as.factor(data.loc$Strain)
    data.loc$Test <- as.factor(data.loc$Test)
    data.loc$RepNo <- as.factor(data.loc$RepNo)
    data.loc$Row <- as.factor(data.loc$Row)
    data.loc$Range <- as.factor(data.loc$Range)
    Tsts <- levels(factor(data.loc$Test))
    data.loc <- droplevels(data.loc)
 
```

Data are filtered for non-missing values, and the analysis loops over locations.

---

## Model Formula

For **Yield** in the first location:

```{r, eval=FALSE}
 if(trait== "Yield" & nLoc==1){
      fixedEqn <- paste(trait," ~ Test + RepNo:Test",sep="")
    }else{ 
     
      fixedEqn <- paste(trait," ~ 1 ",sep="")
    }
    
```

#For other traits and locations:

This minimal intercept model improves h² estimation for traits like Oil and Protein.

---

## Fit the Random Model

Fit the model treating Strain as random:

```{r, eval=FALSE}

count <- 1
    warningMessages <- c()
    modelRnd_Loc <- tryCatch({
      withCallingHandlers({
        asreml(fixed = as.formula(fixedEqn),
               random=~ Strain,
               data = data.loc,workspace="2048mb")

      }, warning = function(w){
        # This function will execute if a warning is issued
        # Capture the warning message 
        warningMessage <- conditionMessage(w)
        print(paste("Warning captured:", warningMessage))
        if(grepl("not converged",warningMessage)){
          warningMessages <<-  warningMessage
        }
        # Invoke the default warning handler (optional)
        invokeRestart("muffleWarning")
      })
    }, error=function(e) {
      print(paste("Error encountered:", e$message))
    })
    
```

Convergence warnings are handled with updates:

```{r, eval=FALSE}

 if(!is.null(warningMessages)){
      while(grepl("not converged", warningMessages[length(warningMessages)]) && count <=5){
        
        tryCatch({
          withCallingHandlers({
            modelRnd_Loc <- update.asreml(modelRnd_Loc)}, warning=function(w) {
              warningMessage <- conditionMessage(w) 
              print(paste("Warning captured:", warningMessage))
              
              if(grepl("not converged",warningMessage)){
                warningMessages <<- warningMessage
              }else{warningMessages <<- "Iterations converged" 
              print(paste(warningMessages[length(warningMessages)]))
              }
            }
          )
        })
        count <- count+1
      }
    }
```

---

## BLUP Predictions

If the model converges:

```{r, eval=FALSE}

 if(trait == "Yield" & nLoc==1){
        BLUPs_Loc <- predict(modelRnd_Loc,classify="Strain",present=c("Strain","Test"))
     
  }else{
        BLUPs_Loc <- predict(modelRnd_Loc,classify="Strain",present=c("Strain"))
  }

```

If multiple tests exist:

```{r, eval=FALSE}

if(length(Tsts)>1){
        if(trait == "Yield" & nLoc==1){
          BLUPs_LocT <- predict(modelRnd_Loc,classify="Strain:Test",present=c("Strain","Test"))
        }else{
          BLUPs_LocT <- predict(modelRnd_Loc,classify="Strain",present=c("Strain"))
        }
}
```

When we use the `predict()` function in ASReml, two important arguments are:

- `classify`
- `present`

They control **which predictions are generated**, and **whether we avoid missing values (NAs)** in the output table.

---

## Wald Tests and Heritability

Obtain Wald tests for fixed effects:

```{r, eval=FALSE}
wald.modelRnd_Loc <- wald.asreml(modelRnd_Loc)
```

Calculate reliability (heritability) estimates:

```{r, eval=FALSE}
i2_Loc <- vpredict(modelRnd_Loc, h2 ~ V1 / (V1 + V2))
```

---

# Part 2: Random-Effects Models with Row/Range

## Model Setup

Additional lists for Row/Range models:

```{r}
modelRnd_Loc_RR_Trts_List <- list()
BLUPs_Loc_RR_Trts_List <- list()
i2_Loc_RR_Trts_List <- list()
wald.modelRnd_Loc_RR_Trts_List <- list()
```

---

## Fixed Effects Model

For **Yield** in the first location:

```{r, eval=FALSE}

      if(trait== "Yield" & nLoc==1){
        fixedEqnRR <- paste(trait," ~ Test + Row + Range + RepNo:Test",sep="")
      }else{ 
        fixedEqnRR <- paste(trait," ~ 1",sep="")
      }
      
```

Otherwise:

```{r, eval=FALSE}
fixedEqnRR <- paste(trait, "~ 1", sep="")
```

---

## Fit the Model

Random model with Strain effect:

```{r, eval=FALSE}
 count <- 1
      warningMessages <- c()
      modelRnd_Loc_RR <- tryCatch({
        withCallingHandlers({
          asreml(fixed = as.formula(fixedEqnRR),
                 random=~ Strain,
                 data = data.loc, workspace="2048mb")
          
        }, warning = function(w){
          # This function will execute if a warning is issued
          # Capture the warning message 
          warningMessage <- conditionMessage(w)
          print(paste("Warning captured:", warningMessage))
          if(grepl("not converged",warningMessage)){
            warningMessages <<-  warningMessage
          }
          # Invoke the default warning handler (optional)
          invokeRestart("muffleWarning")
        })
      }, error=function(e) {
        print(paste("Error encountered:", e$message))
      })
```

Convergence issues are handled via updates.

---
```{r ,eval=FALSE}

 # Check if the warning message contains the specific pattern "1%"
        if(!is.null(warningMessages)){
          while(grepl("not converged", warningMessages[length(warningMessages)]) && count <=5){
            
            tryCatch({
              withCallingHandlers({
                modelRnd_Loc_RR <- update.asreml(modelRnd_Loc_RR)}, warning=function(w) {
                  warningMessage <- conditionMessage(w) 
                  print(paste("Warning captured:", warningMessage))
                  
                  if(grepl("not converged",warningMessage)){
                    warningMessages <<- warningMessage
                  }else{warningMessages <<- "Iterations converged" 
                  print(paste(warningMessages[length(warningMessages)]))
                  }
                }
              )
            })
            count <- count+1
          }
        }

```


## BLUP Predictions

Predictions for Row/Range model:

```{r, eval=FALSE}
  if(trait== "Yield" & nLoc==1){
        BLUPs_Loc_RR <-predict(modelRnd_Loc_RR,classify="Strain",present=c("Strain","Test","Row","Range"))
  }else{ 
          BLUPs_Loc_RR <-predict(modelRnd_Loc_RR,classify="Strain",present=c("Strain"))
  }

  
   if(length(Tsts)>1){
          if(trait== "Yield" & nLoc==1){
           BLUPs_LocT_RR <- predict(modelRnd_Loc_RR,classify="Strain:Test",present=c("Strain","Test","Row","Range"))
          }else{ 
            BLUPs_LocT_RR <-predict(modelRnd_Loc_RR,classify="Strain",present=c("Strain"))
          }
   }
 
```

Or simply by Strain if Row/Range are excluded.

---

## Wald Tests and Heritability

Wald tests:

```{r, eval=FALSE}
wald.modelRnd_Loc_RR <- wald.asreml(modelRnd_Loc_RR)
```

Heritability estimates:

```{r, eval=FALSE}
i2_Loc_RR <- vpredict(modelRnd_Loc_RR, h2 ~ V1 / (V1 + V2))
```

---

# Part 3: Model Comparison with AIC

For each trait, AIC values are compared:

```{r, eval=FALSE}
AICTable_Rnd[nLoc, 1] <- summary(modelRnd_Loc_List[[nLoc]])$aic
AICTable_Rnd[nLoc, 2] <- summary(modelRnd_Loc_RR_List[[nLoc]])$aic
```

Models with spatial terms (Row/Range) are evaluated against simpler random-only models.

---

# Part 4: Summarizing Heritability Estimates

All heritability values across locations and traits are compiled:

```{r, eval=FALSE}
h2_Estimates_Tab <- do.call(cbind.data.frame, lapply(i2_Loc_RR_Trts_List, function(x)
  do.call(rbind.data.frame, lapply(x, function(y) y))))
h2_Estimates_Tab <- apply(h2_Estimates_Tab, 2, function(x) round(x, 3))
write.csv(h2_Estimates_Tab, "PYT24_H2_LocWise.csv", quote = FALSE)
```

This generates a table of location-wise h² estimates for all traits.

---

# Summary

This code section:

✅ fits random models for genotype BLUP estimation  
✅ handles convergence with iterative updates  
✅ fits optional Row/Range spatial models  
✅ calculates location-specific heritability  
✅ stores all outputs for further analysis


