---
title: "PYT 2024 Data Preparation and Quality Control"
author: "Vishnu Ramasubramanian"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document describes the R workflow for preparing and cleaning the 2024 Preliminary Yield Trial (PYT) data. Steps include:

- Reading the raw data
- Filtering and transforming trait data
- Running ANOVA for outlier detection
- Performing quality control
- Saving the final cleaned dataset

---

# Load Data

## Set Working Directory

```{r, eval=FALSE}
setwd("C:/Users/vramasub/Desktop/PYT2024_Analysis/")
```

## Read CSV Data

```{r, eval=FALSE}
All_PYT_2024_V0 <- read.csv("PYTs2024_Data_V2.csv", header = TRUE)
```

Rename the Test column:

```{r, eval=FALSE}
colnames(All_PYT_2024_V0)[which(colnames(All_PYT_2024_V0) %in% "Test..")] <- "Test"
```

Convert Test to a factor and create composite keys:

```{r, eval=FALSE}
All_PYT_2024_V0$Test <- as.factor(All_PYT_2024_V0$Test)
All_PYT_2024_V0$LocTest <- paste(All_PYT_2024_V0$Book.Name, All_PYT_2024_V0$Test, sep = "-")
All_PYT_2024_V0$StrLocTestRep <- paste(All_PYT_2024_V0$Alias.pedigree,
                                       All_PYT_2024_V0$LocTest,
                                       All_PYT_2024_V0$RepNo, sep = "-")
```

---

# Remove Rows with Missing IDs

Identify rows with missing `UniqID`:

```{r, eval=FALSE}
naInd1 <- which(is.na(All_PYT_2024_V0$UniqID))
```

Remove those rows:

```{r, eval=FALSE}
if(length(naInd1) > 0){
  All_PYT_2024_V1 <- All_PYT_2024_V0[-naInd1,]
} else {
  All_PYT_2024_V1 <- All_PYT_2024_V0
}
```

Check dimensions:

```{r, eval=FALSE}
dim(All_PYT_2024_V1)
# [1] 3031 42
```

---

# Cleaning Maturity Rating

Ensure maturity ratings stay within range:

```{r, eval=FALSE}
All_PYT_2024_V1$`Maturity.rating` <- as.numeric(as.character(All_PYT_2024_V1$`Maturity.rating`))
All_PYT_2024_V1$`Maturity.rating`[All_PYT_2024_V1$`Maturity.rating` > 10] <- NA
All_PYT_2024_V2 <- All_PYT_2024_V1[!is.na(All_PYT_2024_V1$`Maturity.rating`), ]
```

---

# Prepare Yield Data

Convert yield column to numeric:

```{r, eval=FALSE}
All_PYT_2024_V2$Yield_bu_acre_13 <- as.numeric(All_PYT_2024_V2$Yield_bu_acre_13)
```

Assign to working data:

```{r, eval=FALSE}
All_PYT_2024 <- All_PYT_2024_V2
```

---

# Define Traits and Formulas

List of traits:

```{r}
traits <- c("Yield_bu_acre_13", "Oil", "Protein", "Maturity.rating")
```

Define model formulas:

```{r}
formulas <- list(
  Yield_bu_acre_13 = Yield_bu_acre_13 ~ Book.Name + RepNo:Book.Name + Alias.pedigree,
  Oil = Oil ~ Book.Name + Alias.pedigree,
  Protein = Protein ~ Book.Name + Alias.pedigree,
  Maturity.rating = Maturity.rating ~ Book.Name + RepNo:Book.Name + Alias.pedigree
)
```

Alternative formulas with interactions:

```{r}
formulas_GL <- list(
  Yield_bu_acre_13 = Yield_bu_acre_13 ~ Book.Name * Alias.pedigree + RepNo:Book.Name,
  Oil = Oil ~ Book.Name * Alias.pedigree,
  Protein = Protein ~ Book.Name * Alias.pedigree,
  Maturity.rating = Maturity.rating ~ Book.Name * Alias.pedigree + RepNo:Book.Name
)
```

---

# Function to Fit ANOVA and Remove Outliers

```{r, eval=FALSE}
process_trait <- function(trait, formula, data) {
  aov_model <- aov(formula, data = data)
  predictions <- fitted(aov_model)
  residuals <- residuals(aov_model)
  rmse <- sqrt(mean(residuals^2, na.rm = TRUE))
  std_residuals <- residuals / rmse

  outlier_indices <- which(abs(std_residuals) > 2.5)

  if(length(aov_model$na.action) > 0){
    data.filt <- data[-aov_model$na.action,]
  } else {
    data.filt <- data
  }

  filtered_data <- data.filt[-outlier_indices, ]
  filtered_data[paste0(trait, "_Pred")] <- predictions[-outlier_indices]
  filtered_data[paste0(trait, "_StdResid")] <- std_residuals[-outlier_indices]

  return(filtered_data)
}
```

---

# Process Each Trait

```{r, eval=FALSE}
results_list <- lapply(traits, function(trait) {
  process_trait(trait, formulas[[trait]], All_PYT_2024)
})
```

---

# Prepare Filtered Data

Select relevant columns:

```{r, eval=FALSE}
results_list_filt <- lapply(seq_along(results_list), function(x) {
  if(traits[x] == "Yield_bu_acre_13") {
    yldInd <- which(colnames(results_list[[x]]) %in% c("Yield_bu_acre_13","StrLocTestRep"))
    selCols <- c(1:yldInd[1], yldInd[2])
    results_list[[x]][, selCols]
  } else {
    trtInd <- which(colnames(results_list[[x]]) %in% c(traits[x],"StrLocTestRep"))
    results_list[[x]][, trtInd]
  }
})
```

Merge all trait data:

```{r, eval=FALSE}
All_PYT_2024_Trts <- Reduce(function(x, y) merge(x, y, by = "StrLocTestRep", all = TRUE), results_list_filt)
dim(All_PYT_2024_Trts)
```

---

# Quality Control

Remove “Bad” plots:

```{r, eval=FALSE}
All_PYT_2024_Trts_QC <- subset(All_PYT_2024_Trts, Plot.Status != "Bad")
dim(All_PYT_2024_Trts_QC)
```

Summarize traits:

```{r, eval=FALSE}
summary(All_PYT_2024_Trts_QC$Yield_bu_acre_13)
summary(All_PYT_2024_Trts_QC$Oil)
summary(All_PYT_2024_Trts_QC$Protein)
```

Identify low-yielding lines:

```{r, eval=FALSE}
yldlt10 <- which(All_PYT_2024_Trts_QC$Yield_bu_acre_13 <= 10)
table(All_PYT_2024_Trts_QC[yldlt10, "Entry.Book.Name"])
```

Remove problematic trials:

```{r, eval=FALSE}
Tst16WAInd <- which(All_PYT_2024_Trts_QC$Test == "16" &
                    All_PYT_2024_Trts_QC$Book.Name == "WA")

rmQC2Ind <- union(Tst16WAInd, yldlt10)
All_PYT_2024_Trts_QC_V2 <- All_PYT_2024_Trts_QC[-rmQC2Ind, ]
dim(All_PYT_2024_Trts_QC_V2)
table(All_PYT_2024_Trts_QC_V2$Book.Name)
```

---

# Save Cleaned Data

Export final cleaned data:

```{r, eval=FALSE}
write.csv(All_PYT_2024_Trts_QC_V2, "PYT_2024_All_Traits_QC_V2.csv",
          quote = FALSE, row.names = FALSE)
```

---

# Conclusion

This pipeline ensures a robust, cleaned dataset for downstream analysis. The final data set is free of outliers and low-quality records, ready for statistical modeling and reporting.


