---
title: "Demo for BLUEs Estimation with ASReml Models including Row & Range Effects with PYT2024 Data"
author: "Vishnu Ramasubramanian"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This document explains an R script for fitting **ASReml models with Row and Range effects**. The script is intended for analyzing multi-location, multi-trait field trial data in plant breeding experiments.

---

# Purpose of the Code

The purpose of this code is to fit location-specific ASReml linear models for multiple traits. These models:

- account for **row and range effects** (i.e. spatial effects) in the field layout
- estimate BLUEs (Best Linear Unbiased Estimates) for genotypes
- optionally estimate strain means per test
- perform Wald tests to evaluate significance of fixed effects

Row and Range capture systematic spatial variability in field trials, improving the precision of estimates for genotype performance.

---

# Detailed Walkthrough of the Code

---

## Initialize Lists

We start by initializing lists to store results for each trait:

```{r}
model_Loc_RR_Trts_List <- list()
BLUEs_Loc_RR_Trts_List <- list()
BLUEs_LocTst_RR_Trts_List <- list()
wald.model_Loc_RR_Trts_List <- list()
```

These lists will hold:

- fitted ASReml models for each location
- BLUE predictions for strains
- strain predictions within tests (if multiple tests exist)
- Wald test results

---

## Loop Over Traits

The script processes each trait separately:

```{r, eval=FALSE}
for(trait in trait.list) {
  # analysis steps follow ...
}
```

---

## Filter Data for the Trait

```{r, eval=FALSE}
trtInd <- which(!is.na(data.regular[,trait]))
data.regular.trait <- data.regular[trtInd,]
data.regular.trait$Location <- as.factor(data.regular.trait$Location)
Locations <- levels(factor(data.regular.trait$Location))
```

- Removes missing values for the trait.
- Converts `Location` to a factor.
- Retrieves all unique locations.

---

## Loop Over Locations

The model is fit separately at each location:

```{r, eval=FALSE}
for(nLoc in 1:length(Locations)) {
  loc <- Locations[nLoc]
  data.loc <- data.regular.trait[data.regular.trait$Location == loc, ]
```

---

## Convert Columns to Factors

```{r, eval=FALSE}
data.loc$Strain <- as.factor(data.loc$Strain)
data.loc$Test <- as.factor(data.loc$Test)
data.loc$RepNo <- as.factor(data.loc$RepNo)
data.loc$Row <- as.factor(data.loc$Row)
data.loc$Range <- as.factor(data.loc$Range)
data.loc <- droplevels(data.loc)
```

Ensures categorical variables are treated as factors for ASReml.

---

## Construct the Model Formula

Two different model formulas are used:

- For **Yield** and **Maturity.rating**:

    $$ \text{Trait} \sim \text{Test} + \text{Row} + \text{Range} + \text{Strain} + \text{RepNo}:\text{Test} $$

- For other traits (e.g. Oil, Protein):

    $$ \text{Trait} \sim \text{Strain} $$

### Code:

```{r, eval=FALSE}
if(trait == "Yield" | trait == "Maturity.rating") {
  fixedEqnRR <- paste(trait, "~ Test + Row + Range + Strain + RepNo:Test", sep="")
} else {
  fixedEqnRR <- paste(trait, "~ Strain", sep="")
  rowPresent <- FALSE
  testPresent <- FALSE
}
```

---

## Fit the ASReml Model

We fit the model using ASReml, capturing warnings:

```{r, eval=FALSE}
model_Loc_RR <- tryCatch({
  withCallingHandlers({
    asreml(fixed = as.formula(fixedEqnRR),
           data = data.loc,
           workspace = "2048mb", ai.sing = TRUE)
  }, warning = function(w) {
    warningMessage <- conditionMessage(w)
    print(paste("Warning captured:", warningMessage))
    if (grepl("not converged", warningMessage)) {
      warningMessages <<- warningMessage
    }
    invokeRestart("muffleWarning")
  })
}, error = function(e) {
  print(paste("Error encountered:", e$message))
})
```

---

## Handle Convergence Warnings

If convergence fails, the model is updated up to 5 times:

```{r, eval=FALSE}
if (!is.null(warningMessages)) {
  while (grepl("not converged", warningMessages[length(warningMessages)]) && count <= 5) {
    tryCatch({
      withCallingHandlers({
        model_Loc_RR <- update.asreml(model_Loc_RR)
      }, warning = function(w) {
        warningMessage <- conditionMessage(w)
        print(paste("Warning captured:", warningMessage))
        if (grepl("not converged", warningMessage)) {
          warningMessages <<- warningMessage
        } else {
          warningMessages <<- "Iterations converged"
          print(paste(warningMessages[length(warningMessages)]))
        }
      })
    })
    count <- count + 1
  }
}
```

---

## Handle Singularities

If errors due to singularities arise (e.g. when Row/Range effects are confounded), the model is refitted without Row/Range terms:

```{r, eval=FALSE}
if (!is.null(errorMessages)) {
  if (grepl("singularities", errorMessages[length(errorMessages)])) {
    fixedEqnRR <- paste(trait, "~ Test + Strain + RepNo:Test", sep="")
    rowPresent <- FALSE
    model_Loc_RR <- tryCatch({
      asreml(fixed = as.formula(fixedEqnRR),
             data = data.loc,
             workspace = "2048mb", ai.sing = TRUE)
    }, warning = function(w) {
      # handle warnings
    }, error = function(e) {
      # handle errors
    })
  }
}
```

---

## Predict BLUEs

If the model fits successfully, predictions are obtained:

```{r, eval=FALSE}
BLUEs_Loc_RR <- tryCatch({
  withCallingHandlers({
    if(rowPresent & testPresent) {
      predict(model_Loc_RR, classify = "Strain",
              present = c("Strain", "Test", "Row", "Range"),
              workspace = "2048mb")
    } else if (!rowPresent & testPresent) {
      predict(model_Loc_RR, classify = "Strain",
              present = c("Strain", "Test"),
              workspace = "2048mb")
    } else {
      predict(model_Loc_RR, classify = "Strain",
              present = c("Strain"),
              workspace = "2048mb")
    }
  }, warning = function(w) {
    warningMessage <- conditionMessage(w)
    print(paste("Warning captured:", warningMessage))
    invokeRestart("muffleWarning")
  })
}, error = function(e) {
  print(paste("Error encountered:", e$message))
})
```


When we use the `predict()` function in ASReml, two important arguments are:

- `classify`
- `present`

They control **which predictions are generated**, and **whether we avoid missing values (NAs)** in the output table.

---

## Predict BLUEs by Strain × Test

If there are multiple tests:

```{r, eval=FALSE}
if (length(Tsts) > 1) {
  BLUEs_LocT_RR <- tryCatch({
    predict(model_Loc_RR, classify = "Strain:Test",
            present = c("Strain", "Test", "Row", "Range"),
            workspace = "2048mb")
  }, warning = function(w) {
    # handle warnings
  }, error = function(e) {
    # handle errors
  })
} else {
  BLUEs_LocT_RR <- NULL
}
```

---

## Wald Tests

Wald tests are performed to test significance of fixed effects:

```{r, eval=FALSE}
wald.model_Loc_RR <- tryCatch({
  wald.asreml(model_Loc_RR)
}, warning = function(w) {
  warningMessage <- conditionMessage(w)
  print(paste("Warning captured:", warningMessage))
  invokeRestart("muffleWarning")
}, error = function(e) {
  print(paste("Error encountered:", e$message))
})
```

---

## Save Results

Results are stored in lists:

```{r, eval=FALSE}
model_Loc_RR_List[[nLoc]] <- model_Loc_RR
BLUEs_Loc_RR_List[[nLoc]] <- BLUEs_Loc_RR
BLUEs_LocTst_RR_List[[nLoc]] <- BLUEs_LocT_RR
wald.model_Loc_RR_List[[nLoc]] <- wald.model_Loc_RR
```

And finally stored for each trait:

```{r, eval=FALSE}
model_Loc_RR_Trts_List[[trait]] <- model_Loc_RR_List
BLUEs_Loc_RR_Trts_List[[trait]] <- BLUEs_Loc_RR_List
BLUEs_LocTst_RR_Trts_List[[trait]] <- BLUEs_LocTst_RR_List
wald.model_Loc_RR_Trts_List[[trait]] <- wald.model_Loc_RR_List
```

---

# Summary

This code fits separate ASReml models for each trait and location, including optional spatial effects (Row and Range) and nested replicate effects. It:

✅ Fits robust models for each location  
✅ Handles convergence warnings  
✅ Detects and corrects singularities  
✅ Predicts BLUEs for strains and strain × test combinations  
✅ Stores results in a structured way

